% ============================================================================
% gtopt: High-Performance Generation and Transmission Expansion Planning
%
% Target journal: IEEE Transactions on Power Systems
% Template: IEEEtran (journal mode)
%
% Overleaf-compatible: upload this entire directory as a project.
% To compile locally: latexmk -pdf main.tex
% ============================================================================

\documentclass[journal]{IEEEtran}

% ---- Encoding ----
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% ---- Math ----
\usepackage{amsmath,amssymb,amsfonts}

% ---- Graphics and tables ----
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{subcaption}

% ---- Typography ----
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[binary-units]{siunitx}

% ---- Code listings ----
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny\color{gray},
  xleftmargin=2em,
  framexleftmargin=1.5em,
}

% ---- Hyperlinks (load last among standard packages) ----
\usepackage[hidelinks,colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{cite}

% ---- TikZ ----
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,shapes.geometric,fit,calc,patterns,decorations.pathreplacing}

% ---- Algorithms ----
\usepackage{algorithmic}
\usepackage{algorithm}

% ---- Custom commands ----
\newcommand{\gtopt}{\texttt{gtopt}}
\newcommand{\fesop}{FESOP}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\setS}{\mathcal{S}}
\newcommand{\setT}{\mathcal{T}}
\newcommand{\setB}{\mathcal{B}}
\newcommand{\setN}{\mathcal{N}}
\newcommand{\setG}{\mathcal{G}}
\newcommand{\setD}{\mathcal{D}}
\newcommand{\setL}{\mathcal{L}}
\newcommand{\setE}{\mathcal{E}}
\newcommand{\setJ}{\mathcal{J}}
\newcommand{\setR}{\mathcal{R}}
\newcommand{\setW}{\mathcal{W}}
\newcommand{\setH}{\mathcal{H}}
\newcommand{\setZ}{\mathcal{Z}}
\newcommand{\setFR}{\mathcal{F\!R}} % flow rights
\newcommand{\setVR}{\mathcal{V\!R}} % volume rights
\newcommand{\setHP}{\mathcal{H}_P}   % hydraulic pumps
\newcommand{\setLNG}{\mathcal{LNG}}  % LNG terminals
\newcommand{\setUC}{\mathcal{UC}}    % UC generators
\newcommand{\setIZ}{\mathcal{IZ}}    % inertia zones
\newcommand{\setIP}{\mathcal{IP}}    % inertia provisions

% ---- Graphics path ----
\graphicspath{{figures/}{diagrams/}}

% ============================================================================
\begin{document}

\title{%
  \gtopt{}: A Novel High-Performance Tool for\\
  Generation and Transmission Expansion Planning%
}

\author{%
  Marcelo~Matus~A.,~\IEEEmembership{Member,~IEEE,}
  Mar\'ia~del~Pilar~Buitrago~Villada,
  Erick~Sierra~Baeza,
  Carlos~Enrique~Garc\'ia~Bujanda,
  Marcos~Falcone~Salazar,
  and~Gonzalo~Bustos~Turu%
  \thanks{Manuscript received MONTH DAY, 2026; revised MONTH DAY, 2026.}%
  \thanks{M.~Matus~A. is with Black Bear Engineering SpA, Santiago,
    Chile, and the Centro Nacional de Inteligencia Artificial
    (CENIA, FB210017), Santiago, Chile
    (e-mail: marcelo.matus@blackbear.cl).}%
  \thanks{M.~P.~Buitrago~Villada is with the Department of Electrical
    Energy and Automation, Universidad Nacional de Colombia,
    Medell\'in, Colombia.}%
  \thanks{E.~Sierra~Baeza is with Black Bear Engineering SpA, Santiago,
    Chile, and CENIA.}%
  \thanks{C.~E.~Garc\'ia~Bujanda is with Black Bear Engineering SpA,
    Santiago, Chile.}%
  \thanks{M.~Falcone~Salazar is with the School of Engineering,
    Universidad Adolfo Ib\'a\~nez, Santiago, Chile, and formerly with
    Colb\'un~S.A., Santiago, Chile.}%
  \thanks{G.~Bustos~Turu is with the Department of Chemical Engineering,
    Imperial College London, London, UK, and the Centre for Energy,
    Universidad de Chile, Santiago, Chile.}%
  % \thanks{Digital Object Identifier: 10.1109/TPWRS.20XX.XXXXXXX}%
}

% --- Create title and author block ---
\markboth{IEEE Transactions on Power Systems}%
  {Matus~A. \MakeLowercase{\textit{et al.}}: \gtopt{} for GTEP}

\maketitle

% ========== Abstract ==========
\begin{abstract}
The increasing penetration of variable renewable energy and the growing
complexity of multi-sector energy systems demand planning tools that can
co-optimize generation, transmission, and storage investments while
capturing detailed operational constraints over long horizons under
uncertainty.  Existing open-source tools, primarily implemented in Python
and Julia, can face computational bottlenecks when assembling sparse LP/MIP
problems with millions of variables.  This paper presents \gtopt{}, a
novel, high-performance Generation and Transmission Expansion Planning
(GTEP) tool implemented in modern C++26.  The joint investment and
operational planning problem is formulated as a sparse linear program over
a multi-stage, multi-scenario time hierarchy, supporting DC optimal power
flow with Kirchhoff voltage-law constraints, battery energy storage with
charge/discharge efficiency modeling, hydro cascade systems with reservoirs
and turbines, spinning reserve requirements, water rights and irrigation
agreement constraints with volume-dependent dynamic bounds, and capacity
expansion with annualized investment costs.  Three solution methods are
provided: a
monolithic LP solver, Stochastic Dual Dynamic Programming (SDDP) with
multi-scenario apertures, and a novel multi-level cascade decomposition.
Pluggable LP solver backends (CLP, CBC, CPLEX, HiGHS) are loaded at
runtime as dynamic libraries.  Validation against IEEE benchmark networks
(4--57 buses, 13 test cases) and cross-comparison with pandapower DC OPF
demonstrate agreement to solver tolerance ($<10^{-6}$ relative error).
Computational experiments show a 10--27$\times$ speedup in sparse-matrix
assembly over Python-based alternatives, with the largest test case (57-bus,
8760 blocks, ${\sim}18$M variables) assembling in approximately 8 seconds.
\end{abstract}

% ========== Keywords ==========
\begin{IEEEkeywords}
Generation expansion planning, transmission expansion planning, linear
programming, DC optimal power flow, stochastic dual dynamic programming,
battery energy storage, hydro cascade, irrigation agreements, water
rights, sparse LP assembly.
\end{IEEEkeywords}

% ========== Sections ==========
\input{sections/introduction}
\input{sections/literature_review}
\input{sections/formulation}
\input{sections/solution_methods}
\input{sections/architecture}
\input{sections/validation}
\input{sections/case_studies}
\input{sections/conclusions}

% ========== Acknowledgments ==========
% \section*{Acknowledgments}
% Add acknowledgments as needed.

% ========== References ==========
\bibliographystyle{IEEEtran}
\bibliography{references}

% ========== Biographies ==========
\begin{IEEEbiography}{Marcelo Matus~A.}
  received the Electrical Engineer degree from Pontificia Universidad
  Cat\'olica de Chile and the Ph.D. degree in electrical and computer
  engineering from the University of Arizona, Tucson, AZ, USA.  He is
  currently a Senior Researcher at Black Bear Engineering SpA and at
  the Centro Nacional de Inteligencia Artificial (CENIA), Santiago,
  Chile.  He was formerly Subdirector of the Centre for Energy (Centro
  de Energ\'ia), Universidad de Chile.  His research interests include
  power system optimization, mathematical programming, and
  high-performance computing applied to energy systems.  He is the
  lead developer of gtopt and the FESOP framework.
\end{IEEEbiography}

\begin{IEEEbiography}{Mar\'ia del Pilar Buitrago Villada}
  received the B.Sc. and M.Sc. degrees in electrical engineering from
  Universidad Nacional de Colombia, Medell\'in, and the Ph.D. degree in
  engineering from the same university.  Her research focuses on power
  system expansion planning, renewable energy integration, and
  optimization under uncertainty.  She is a co-developer of the FESOP
  framework and has contributed to its application in long-term planning
  of the Colombian and Chilean power systems.
\end{IEEEbiography}

\begin{IEEEbiography}{Erick Sierra Baeza}
  received the Civil Electrical Engineer degree from Universidad de Chile.
  He is currently with Black Bear Engineering SpA, Santiago, Chile, and
  affiliated with CENIA.  He previously worked at the Centre for Energy,
  Universidad de Chile.  His professional experience spans wind energy
  integration, capacity expansion planning, hydrological modeling, and
  development of optimization tools for energy market analysis.
\end{IEEEbiography}

\begin{IEEEbiography}{Carlos Enrique Garc\'ia Bujanda}
  is a Researcher at Black Bear Engineering SpA and affiliated with the
  Department of Informatics, Universidad de Santiago de Chile.  His
  research interests include power systems optimization, renewable energy
  integration planning, and development of computational tools for
  long-term electricity system expansion.  He is a co-developer of the
  FESOP framework.
\end{IEEEbiography}

\begin{IEEEbiography}{Marcos Falcone Salazar}
  received the Electrical Engineer and M.Sc. degrees in electrical
  engineering from Universidad de Chile and the MBA degree from
  Universidad Adolfo Ib\'a\~nez, Santiago, Chile.  He held management
  positions at Colb\'un~S.A. including Sub-manager of Planning and
  Development, where he directed an R\&D initiative between Colb\'un,
  Universidad de Chile, Fondecyt, and CORFO that led to the development
  of a novel electricity market planning and operation model.  He is
  currently a Professor at the School of Engineering, Universidad Adolfo
  Ib\'a\~nez.  His professional interests include power system expansion
  planning, energy project valuation, and electricity market design.
\end{IEEEbiography}

\begin{IEEEbiography}{Gonzalo Bustos Turu}
  received the B.Sc. and M.Sc. degrees in electrical engineering from
  Universidad de Chile in 2009, the M.Sc. degree in sustainable energy
  futures from Imperial College London in 2013, and the Ph.D. degree
  from Imperial College London in 2018, working on smart urban energy
  systems under the supervision of Prof.~Nilay Shah.  He is currently
  a Research Associate at Imperial College London and affiliated with
  the Centre for Energy, Universidad de Chile.  His research interests
  include urban energy system modeling, demand response, electric vehicle
  integration, and district-level predictive control.
\end{IEEEbiography}

\end{document}
