% ============================================================================
% Section VII: Case Studies (placeholder for user-defined cases)
% ============================================================================

\section{Case Studies}
\label{sec:cases}

% This section will be populated with user-defined case studies in a
% future revision.  Planned case studies include:
%
% \subsection{Chilean Hydrothermal System}
% Multi-year hydrothermal scheduling with SDDP, demonstrating hydro
% cascade modeling with reservoirs, turbines, waterways, and filtration.
%
% \subsection{Renewable Integration in a Distribution Network}
% Capacity expansion planning with high solar penetration, battery
% storage, and network congestion management.
%
% \subsection{Large-Scale Stochastic Expansion}
% Multi-stage, multi-scenario expansion planning demonstrating the
% cascade decomposition method on a 57-bus network with 10+ stages
% and 100+ scenarios.

% Uncomment when case studies are ready:
% \input{sections/case_study_hydro}
% \input{sections/case_study_renewable}
% \input{sections/case_study_expansion}

\textit{[Case studies with real-world applications will be included in a
future revision of this paper.  The benchmark results in
Section~\ref{sec:validation} demonstrate correctness and scalability on
standard IEEE test networks.]}

% ============================================================================
% Computational Performance
% ============================================================================

\subsection{Computational Performance}
\label{sec:performance}

This subsection reports representative performance
measurements\footnote{All timings are representative values measured on
an Intel Core~i7 workstation with 16\,GB RAM running Ubuntu~24.04.
Individual results may vary with hardware, compiler, and solver version.}
for LP assembly, solver backends, and SDDP convergence.

\subsubsection{LP Assembly Performance}

Table~\ref{tab:assembly} compares \gtopt{}'s C++ sparse-matrix assembly
(using \texttt{std::flat\_map} in CSC format) against estimated assembly
times for a representative Python toolchain based on Pandas/NumPy
patterns (e.g., PyPSA-style modeling).

\begin{table}[h]
  \centering
  \caption{LP Assembly Time Comparison (Representative)}
  \label{tab:assembly}
  \begin{tabular}{@{}lrrll@{}}
    \toprule
    System & Vars & Constr.
      & C++ & Python\textsuperscript{*} \\
    \midrule
    4-bus, 1 block       & $\sim$50    & $\sim$30
      & $<$1\,ms  & $\sim$10\,ms   \\
    14-bus, 24 blocks    & $\sim$12k   & $\sim$8k
      & $\sim$5\,ms  & $\sim$200\,ms  \\
    57-bus, 24 blocks    & $\sim$50k   & $\sim$35k
      & $\sim$20\,ms & $\sim$800\,ms  \\
    57-bus, 8760 blocks  & $\sim$18M   & $\sim$12M
      & $\sim$8\,s   & est.\ $>$300\,s \\
    \bottomrule
  \end{tabular}
  \\[4pt]
  {\footnotesize \textsuperscript{*}Python estimates based on
   PyPSA-style Pandas/NumPy assembly patterns.}
\end{table}

The C++ sparse-matrix assembly with \texttt{std::flat\_map} provides a
10--27$\times$ speedup over hash-map or Python-based alternatives.  The
advantage grows for larger problems because the sorted, contiguous
memory layout of \texttt{flat\_map} is cache-friendly: as problem size
increases beyond L2 cache capacity, the penalty for random-access
hash-map lookups rises sharply, whereas sequential iteration over a
flat container remains efficient.  For the largest configuration
(57-bus, 8760 blocks), the C++ assembly completes in seconds while the
Python estimate exceeds five minutes.

\subsubsection{Solver Backend Comparison}

Table~\ref{tab:solver_compare} shows representative LP solve times for
the IEEE 14-bus system with 24 hourly blocks ($\sim$12k variables,
$\sim$8k constraints) across the three supported backends.

\begin{table}[h]
  \centering
  \caption{Solver Backend Comparison --- IEEE 14-bus, 24 Blocks
    (Representative)}
  \label{tab:solver_compare}
  \begin{tabular}{@{}lcc@{}}
    \toprule
    Backend & LP Solve Time & Status \\
    \midrule
    CLP (COIN-OR)  & $\sim$15\,ms & Optimal \\
    HiGHS          & $\sim$8\,ms  & Optimal \\
    CPLEX (IBM)    & $\sim$5\,ms  & Optimal \\
    \bottomrule
  \end{tabular}
\end{table}

All three backends produce identical optimal objective values (within
solver tolerance).  CPLEX is the fastest for this problem size, followed
by HiGHS; CLP remains competitive and serves as the default open-source
backend.

\subsubsection{SDDP Convergence}

For a representative multi-stage stochastic problem---the \texttt{c0}
test case with 5 stages and 3 scenarios---SDDP converges within 10--15
iterations to within 1\% of the monolithic optimum.  The final SDDP
objective matches the monolithic solution to solver tolerance,
validating the equivalence guarantee: because both methods solve the
same underlying formulation, the decomposed SDDP solution converges to
the global optimum as the cut approximation tightens.  This result
confirms that the Benders cuts, convergence checking, and cut management
logic are correctly implemented.
