% ============================================================================
% Section I: Introduction
% ============================================================================

\section{Introduction}
\label{sec:introduction}

The global energy transition toward low-carbon power systems requires
sophisticated planning tools capable of co-optimizing generation,
transmission, and storage investments while capturing the operational
complexity of variable renewable energy sources and hydro
cascades~\cite{lumbreras2016progressive}.  Generation and Transmission
Expansion Planning (GTEP) is the class of optimization problems that
determines the minimum-cost combination of infrastructure investments
(CAPEX) and dispatch decisions (OPEX) over long planning horizons under
uncertainty~\cite{hemmati2013comprehensive, mahdavi2019tep_survey}.

Traditional GTEP tools---both commercial (PLEXOS, PLP/PROMAX) and
academic---have been implemented in Fortran or commercial modeling
languages, limiting accessibility and extensibility.  The emergence of
open-source alternatives such as PyPSA~\cite{brown2018pypsa},
GenX~\cite{jenkins2017genx}, and IDAES-GTEP~\cite{idaes_gtep2024} has
broadened access to planning tools through Python and Julia
implementations.  However, these interpreted-language tools face
computational bottlenecks when assembling large sparse LP/MIP problems
with millions of variables and constraints~\cite{frank2016or_opf}.
A tradeoff exists between the accessibility of high-level modeling
languages and the computational efficiency required for large-scale
stochastic expansion planning with detailed network constraints.

This paper presents \gtopt{}, a novel high-performance GTEP tool
implemented in modern C++26 that bridges this gap by combining the
computational efficiency of compiled-language sparse-matrix assembly
with the accessibility of JSON/Parquet-based data
interfaces.  \gtopt{} descends from the FESOP (Fabulous Energy System
Optimizer) framework~\cite{buitrago2022fesop}, which extended long-term
hydrothermal scheduling tools~\cite{pereira2016hydrothermal} to include
capacity expansion, variable renewable energy representation, and modern
flexibility mechanisms.

The main contributions of this work are:

\begin{enumerate}
  \item A complete mathematical formulation of the GTEP problem as a
    sparse LP/MIP, including DC optimal power flow with Kirchhoff
    constraints, battery storage with charge/discharge efficiencies,
    hydro cascades with reservoirs and turbines, spinning reserves,
    a generic water rights layer for irrigation agreement modeling,
    and multi-stage capacity expansion
    (Section~\ref{sec:formulation}).

  \item Three solution methods---monolithic LP, SDDP with multi-scenario
    apertures, and a novel multi-level cascade decomposition---with
    formal equivalence guarantees
    (Section~\ref{sec:methods}).

  \item A modular software architecture with pluggable solver backends
    loaded as dynamic libraries, native Apache Parquet I/O, and a
    comprehensive Python toolchain for data preparation and
    cross-validation (Section~\ref{sec:architecture}).

  \item Validation against 13 IEEE benchmark cases (4--57 buses)
    including battery storage and capacity expansion variants, with
    cross-validation against pandapower DC OPF demonstrating agreement
    to solver tolerance (Section~\ref{sec:validation}).
\end{enumerate}

The remainder of this paper is organized as follows.
Section~\ref{sec:literature} reviews the state of the art in GTEP tools
and methods.  Section~\ref{sec:formulation} presents the mathematical
formulation.  Section~\ref{sec:methods} describes the three solution
methods.  Section~\ref{sec:architecture} details the software architecture.
Section~\ref{sec:validation} presents validation results on IEEE benchmark
networks.  Section~\ref{sec:conclusions} concludes with a summary and
future research directions.
