% ============================================================================
% Section IV: Solution Methods
% ============================================================================

\section{Solution Methods}
\label{sec:methods}

\gtopt{} implements three complementary solution methods, selectable via the
\texttt{method} option.  All three solve the same underlying GTEP formulation
(Section~\ref{sec:formulation}); they differ in how they decompose the
problem across the time and scenario dimensions.

\subsection{Monolithic LP}
\label{sec:monolithic}

The default method assembles the entire multi-scenario, multi-stage problem
into a single sparse LP/MIP and passes it to the solver backend.  The LP
is constructed in compressed sparse column (CSC) format via the
\texttt{FlatLinearProblem} class.

For a system with $N$ buses, $G$ generators, $D$ demands, $L$ lines,
$E$ batteries, $B$ blocks per stage, $T$ stages, and $S$ scenarios, the
approximate problem size is:

\begin{table}[h]
  \centering
  \caption{LP Problem Size Estimates}
  \label{tab:lp_size}
  \begin{tabular}{@{}ll@{}}
    \toprule
    Quantity & Approximate count \\
    \midrule
    Operational variables & $(G{+}2D{+}2L{+}3E{+}N) \times S \times T \times B$ \\
    Investment variables  & $(G{+}D{+}L{+}E) \times T$ \\
    Bus balance rows      & $N \times S \times T \times B$ \\
    Kirchhoff rows        & $L \times S \times T \times B$ (if DC OPF) \\
    SoC balance rows      & $E \times S \times T \times B$ \\
    Capacity rows         & $(G{+}D{+}L{+}E) \times T$ \\
    \bottomrule
  \end{tabular}
\end{table}

The monolithic method is exact (no decomposition error) and is recommended
for problems where the total number of LP variables fits in memory.  For
the IEEE~57-bus system with 24 hourly blocks, the LP has approximately
$57 \times 24 \times (7 + 84 + 160 + 57) \approx 421\,000$ variables.

\subsection{Stochastic Dual Dynamic Programming (SDDP)}
\label{sec:sddp}

For multi-stage stochastic problems with many scenarios, \gtopt{} implements
SDDP following the algorithm of Pereira and
Pinto~\cite{pereira1991sddp}.  The stages are decomposed into
individual subproblems connected by Benders optimality cuts that
approximate the expected future cost.

\begin{algorithm}[h]
  \caption{SDDP Algorithm in \gtopt{}}
  \label{alg:sddp}
  \begin{algorithmic}[1]
    \STATE Initialize: empty cut sets $\mathcal{C}_t = \emptyset$
      for all $t$
    \FOR{iteration $k = 1, 2, \ldots, K$}
      \STATE \textbf{Forward pass}: Solve stages $t=1,\ldots,T$
        sequentially for sampled scenarios; record state variables
        (SoC, reservoir volumes)
      \STATE \textbf{Backward pass}: For $t=T,\ldots,1$, solve
        each stage for all aperture scenarios; compute dual
        variables; generate Benders cuts:
        \[
          \alpha_t \geq \hat{\alpha}_t + \hat{\pi}_t^\top
            (x_t - \hat{x}_t)
        \]
      \STATE Add cuts to $\mathcal{C}_t$
      \STATE Check convergence: upper/lower bound gap $\leq \epsilon$
    \ENDFOR
  \end{algorithmic}
\end{algorithm}

\paragraph{Benders Optimality Cuts.}
The core mechanism of SDDP is the Benders optimality cut, which
approximates the expected cost-to-go function as a piecewise-linear
lower bound.  At each backward step for stage~$t$, the cut takes the
form:
\begin{equation}
  \label{eq:benders_cut}
  \alpha_t \;\geq\; \hat{z}_{t+1}
    + \sum_{i} \hat{\mu}_i \bigl(x_i - \hat{x}_i\bigr),
\end{equation}
where $\alpha_t$ is the future cost variable (the cost-to-go
approximation at stage~$t$), $\hat{z}_{t+1}$ is the optimal objective
value of the stage-$(t{+}1)$ subproblem evaluated at the trial point,
$\hat{\mu}_i$ are the dual multipliers of the state-linking constraints
(coupling successive stages through reservoir volumes, state-of-charge,
etc.), and $\hat{x}_i$ are the trial state variable values from the
forward pass.  As iterations proceed, the accumulated cuts progressively
tighten the piecewise-linear outer approximation of the true expected
cost-to-go, driving the lower bound upward toward the optimum.

\paragraph{Cut Coefficient Modes.}
\gtopt{} supports two modes for computing the cut coefficients
$\hat{\mu}_i$ in~\eqref{eq:benders_cut}.  The \texttt{reduced\_cost}
mode (default) extracts dual information from the reduced costs of the
LP solution, which is numerically robust when state variables are at
their bounds.  The \texttt{row\_dual} mode instead reads the row dual
values of the state-linking constraints directly.  Both modes produce
valid cuts; the choice can affect convergence speed depending on problem
structure and solver backend.

\paragraph{Convergence Criterion.}
Convergence is checked at the end of each iteration by comparing the
upper bound (the average total cost across forward-pass scenario
samples) with the lower bound (the optimal objective of the root-stage
subproblem, which includes all accumulated cuts).  The algorithm
terminates when the relative gap satisfies:
\[
  \frac{z^{\text{upper}} - z^{\text{lower}}}{z^{\text{upper}}}
  \;\leq\; \varepsilon,
\]
where the default tolerance is $\varepsilon = 0.01$ (1\%).

\gtopt{}'s SDDP implementation includes several enhancements:

\begin{itemize}
  \item \textbf{Apertures}: configurable backward-pass scenario sampling
    that groups scenarios into subsets for cut generation, reducing
    computational cost while preserving solution quality.
  \item \textbf{Elastic filters}: infeasible subproblems are handled by
    relaxing state variable bounds with penalty costs, ensuring the
    algorithm does not stall.
  \item \textbf{Boundary cuts}: pre-computed future cost approximations
    at the terminal stage, enabling rolling horizon studies.
  \item \textbf{Hot start}: cuts from previous runs can be loaded to
    accelerate convergence.
  \item \textbf{Cut management}: dominated cuts are periodically purged
    to control LP size growth.
\end{itemize}

\begin{figure}[t]
  \centering
  \begin{tikzpicture}[
    >=Stealth,
    stage/.style={draw, rounded corners=2pt, minimum width=0.9cm,
      minimum height=0.65cm, font=\scriptsize, thick},
    fwd/.style={stage, fill=blue!15},
    bwd/.style={stage, fill=orange!15},
    arr/.style={->, thick},
    stvar/.style={font=\tiny, blue!70},
    cutlbl/.style={font=\tiny, red!60},
  ]
    % Forward pass label
    \node[font=\scriptsize\bfseries, blue!70] at (-1.2,1.8)
      {Forward};

    % Forward pass stages
    \node[fwd] (f1) at (0,1.8) {$t{=}1$};
    \node[fwd] (f2) at (1.6,1.8) {$t{=}2$};
    \node[font=\scriptsize] at (2.8,1.8) {$\cdots$};
    \node[fwd] (fT1) at (3.8,1.8) {$T{-}1$};
    \node[fwd] (fT) at (5.4,1.8) {$t{=}T$};

    % Forward arrows
    \draw[arr, blue!60] (f1) -- (f2)
      node[stvar, above, midway] {$x_1$};
    \draw[arr, blue!60] (f2) -- (2.4,1.8);
    \draw[arr, blue!60] (3.2,1.8) -- (fT1);
    \draw[arr, blue!60] (fT1) -- (fT)
      node[stvar, above, midway] {$x_{T-1}$};

    % State variable annotation
    \node[stvar, above=3pt of f1.north east, anchor=south]
      {SoC, $v_r$};

    % Backward pass label
    \node[font=\scriptsize\bfseries, orange!70] at (-1.2,0)
      {Backward};

    % Backward pass stages
    \node[bwd] (b1) at (0,0) {$t{=}1$};
    \node[bwd] (b2) at (1.6,0) {$t{=}2$};
    \node[font=\scriptsize] at (2.8,0) {$\cdots$};
    \node[bwd] (bT1) at (3.8,0) {$T{-}1$};
    \node[bwd] (bT) at (5.4,0) {$t{=}T$};

    % Backward arrows (right to left)
    \draw[arr, red!50] (bT) -- (bT1)
      node[cutlbl, below, midway] {cut};
    \draw[arr, red!50] (bT1) -- (3.2,0);
    \draw[arr, red!50] (2.4,0) -- (b2);
    \draw[arr, red!50] (b2) -- (b1)
      node[cutlbl, below, midway] {cut};

    % Connecting arrows (forward to backward)
    \draw[->, dashed, gray, thin] (fT.south) -- (bT.north);
    \draw[->, dashed, gray, thin] (f1.south) -- (b1.north);

    % Convergence check
    \node[draw, rounded corners=3pt, fill=gray!10,
      font=\scriptsize, minimum width=2.5cm] (conv) at (2.7,-1.3)
      {$|z^\text{UB} - z^\text{LB}|/z^\text{UB} \leq \varepsilon$?};
    \draw[->, thick, gray] (b1.south) |- (conv.west);

    % Loop back
    \draw[->, thick, gray, dashed]
      (conv.east) -| (5.8,-1.3) -- (5.8,2.2) -- (f1.north)
      node[font=\tiny, pos=0.25, right] {iterate};
  \end{tikzpicture}
  \caption{SDDP forward-backward iteration: the forward pass solves
    stages sequentially passing state variables (battery SoC, reservoir
    volume); the backward pass generates Benders cuts. Convergence is
    checked via the upper/lower bound gap.}
  \label{fig:sddp_flow}
\end{figure}

\subsection{Cascade Decomposition}
\label{sec:cascade}

The cascade method is a novel multi-level hybrid that decomposes the
problem hierarchically:

\begin{enumerate}
  \item \textbf{Level~0 (strategic)}: A coarsened version of the problem
    (aggregated blocks, reduced scenarios) is solved monolithically to
    determine investment decisions and approximate cost-to-go functions.
  \item \textbf{Level~1 (tactical)}: SDDP is applied to the full
    stochastic problem using the Level~0 investments as warm-start and
    the cost-to-go cuts as initial approximations.
  \item \textbf{Level~2+ (operational)}: Additional refinement levels
    can solve detailed operational subproblems with fixed investments.
\end{enumerate}

The cascade method inherits SDDP's convergence guarantees while using the
monolithic coarse solution to dramatically reduce the number of SDDP
iterations required.  It is particularly effective for long-horizon problems
(10+ stages) with many scenarios.

\subsection{Solver Backend Architecture}

All three methods delegate LP/MIP solution to a pluggable solver backend
loaded as a shared library at runtime.  Currently supported backends are:

\begin{itemize}
  \item \textbf{CLP/CBC} (COIN-OR): Default open-source backend.  CLP for
    LP, CBC for MIP~\cite{forrest2005cbc}.
  \item \textbf{CPLEX} (IBM): Commercial high-performance solver, accessed
    via the C Callable Library.
  \item \textbf{HiGHS}: Open-source solver with parallel simplex and
    interior-point methods~\cite{huangfu2018highs}.
\end{itemize}

The backend is auto-detected by priority (CPLEX $>$ HiGHS $>$ CBC $>$ CLP)
or explicitly selected via the \texttt{--solver} flag.  Solver-specific
options (algorithm, threads, tolerances, time limits) are configured
through the \texttt{solver\_options} JSON section.
