% ===========================================================
% abstract.tex — IEEEtran abstract + keywords
% ===========================================================
\begin{abstract}
The decarbonisation of the Chilean power sector and the country's
commitment to a carbon-tax escalation under Law~20.780 require
operational signals that are both \emph{spatially} resolved
(per-bus) and \emph{marginally} attributed (per-unit) — the natural
analogue of a Locational Marginal Emission factor.  At present the
Coordinador El\'ectrico Nacional (CEN) publishes only a 15-minute
nodal marginal cost $\lambda^{\mathrm{CEN}}$ and an annual system
average emission factor; no per-bus, per-quarter marginal CO$_2$
intensity $\varepsilon$ is released.  This paper presents
\texttt{cen2gtopt}, a novel reconstruction methodology that
re-engineers $\varepsilon$ for the \num{29} CEN-published 220/500\,kV
buses at 15-minute resolution from public data alone.  The procedure
combines the SCVIC merit-ladder of declared variable costs, the
``generaci\'on real'' hourly dispatch, the ``instrucciones
operacionales'' price-taker classification, and the Marginal Loss
Factors (MLFs) extracted from the PLEXOS PCP/PID solution
\texttt{.accdb} oracle.  A chained-tolerance plus cumulative-spread
clustering rule separates the \num{29} buses into electrically
coherent islands per quarter; for each island the marginal unit is
identified as the dispatched non-forced generator whose declared CV
is closest to the median bus marginal price.  The methodology is
applied to the operating day 2026-04-22.  Across \num{2494}
(bus, quarter) cells, the reconstructed marginal cost $\hat\lambda$
agrees with $\lambda^{\mathrm{CEN}}$ to a mean absolute error of
\SI{0.170}{\USD\per\MWh}, a 90$^{\mathrm{th}}$ percentile of
\SI{0.492}{\USD\per\MWh} and a maximum of
\SI{2.533}{\USD\per\MWh}; after excluding the \SI{13.7}{\percent} of
cells flagged as \texttt{tier\_uncertainty}, the mean drops to
\SI{0.137}{\USD\per\MWh} and the maximum to
\SI{1.34}{\USD\per\MWh}.  Per-island marginal-unit coherence is
exact (\num{0} of \num{474} islands disagree).  The pipeline runs
in approximately \SI{6}{\second} on a warm cache and is published
as part of the \texttt{gtopt} project.
\end{abstract}

\begin{IEEEkeywords}
Locational marginal emissions, marginal CO$_2$ factor, locational
marginal price, Coordinador El\'ectrico Nacional, SCVIC, marginal
loss factor, merit ladder, public-data reconstruction,
short-run-marginal emissions, Chilean power system.
\end{IEEEkeywords}
