% ===========================================================
% discussion.tex
% ===========================================================
\section{Discussion}
\label{sec:discussion}

The pipeline's strengths are also the source of its limitations:
its dependence on public CEN, SCVIC, and PLEXOS feeds confers
auditability and reproducibility but also exposes it to the
cadence, completeness, and rate-limit policies of the publishing
agency.  Six structural limitations deserve explicit attention.

\subsection{Rate-limit dependency}
The CEN SIP API enforces a soft rate limit
(\SI{8}{requests\per\second} on the documented anonymous-read key)
and intermittently returns HTTP~502 on heavy-fetch endpoints such
as \texttt{cmg-programado-pcp} and the per-bar\_transf
\texttt{cmg-real} fan-out.  The pipeline mitigates this through
parallel fetching with bounded concurrency and an aggressive
parquet cache, but a cold-cache day requires roughly \SI{75}{\second}
of wall time and an active CEN session.  Days with API outages
(observed twice in the cached \num{2024}--\num{2026} archive)
require a manual re-fetch the following day.

\subsection{Hand-curated bus aliases}
The bus-name bridge of Section~\ref{sec:bridge} relies on
\num{29} hand-curated overrides in
\texttt{prepopulate\_cache.py} to reconcile PLEXOS node names
that the heuristic alias generator cannot match unambiguously.
Maintenance burden grows linearly with the catalogue: extending
the pipeline to the full \num{1300}+ CEN-published bus list would
require either an expanded overrides table or a more aggressive
fuzzy-matching strategy (e.g.\ Levenshtein distance with substation
catalogue lookup).  The current overrides table is small enough to
be manually audited but is the single largest accumulator of
``soft'' technical debt in the codebase.

\subsection{SCVIC tier-segmentation wedge}
The \texttt{tier\_uncertainty} flag of Section~\ref{sec:eps}
identifies cells where the picker has selected a CV tier whose
next-up neighbour is more than $\$2$/MWh away — i.e.\ where the
true marginal CV is structurally bounded between the two tiers but
the discretisation forces a single value.  These cells account for
\SI{13.7}{\percent} of the 2026-04-22 sample and contribute most
of the worst-case error.  Mitigation requires either a finer
SCVIC ladder (out of the pipeline's control: the ladder is what
the generator declares) or a within-tier interpolation rule (e.g.\
linear in the cumulative dispatch within the tier band).  Within-tier
interpolation is on the roadmap but introduces its own
identifiability problem: the pipeline does not have access to the
PLEXOS-internal piecewise-linear cost segment that was binding at
the LP solve.

\subsection{Loss-factor decomposition}
The per-bus loss correction is not a proper PJM-style
energy/loss/congestion decomposition: the pipeline does not have
access to the per-unit Power Transfer Distribution Factor (PTDF)
matrix, which CEN does not publish.  The MLF read from the
PLEXOS \texttt{.accdb} approximates the loss component but does
not isolate the congestion component.  As a consequence the
pipeline cannot answer counterfactual questions of the form ``how
would $\hat\lambda$ at bus $b$ change if line $\ell$ were
de-rated?'' — a question that the gtopt SDDP solver
(Section~\ref{sec:conclusions}) is well-equipped to answer.

\subsection{Reservoir-hydro as marginal}
The current pipeline excludes reservoir-hydro plants from the
candidate pool, on the rationale that the published feeds do not
expose the implicit water value at which the LP cleared.  Empirically,
this exclusion costs accuracy on hours when a reservoir hydro is
genuinely interior-marginal — typically the late-evening peak when
neither GNL CCGT nor coal is on the merit margin and the published
$\lambda^{\mathrm{CEN}}$ reflects the hydro shadow price on the
Bellman value function.  The natural mitigation is to import the
implicit water value from the \texttt{cmg-programado-pcp} median
and seed reservoir-hydros into the pool with this value as their
synthetic CV; the pipeline is structurally ready for this
extension, blocked only on stable availability of the
\texttt{cmg-programado-pcp} endpoint (which routinely returns
HTTP~502 on the anonymous-read key tier).

\subsection{Single-pollutant scope}
The current emission factor is CO$_2$ only.  Multi-pollutant
extension (NO$_x$, SO$_2$, PM$_{2.5}$) requires per-unit declared
factors from the Registro de Emisiones y Transferencias de
Contaminantes (RETC); the integration is straightforward
(\texttt{FUEL\_EF} becomes \texttt{POLLUTANT\_EF[fuel][pollutant]})
but is not yet implemented.  When implemented it would close the
last remaining methodological gap relative to the
PJM~LME feed~\cite{PJMLME2024}.
