% ===========================================================
% literature_review.tex
% ===========================================================
\section{Background and Related Work}
\label{sec:related}

\subsection{Spot pricing and locational marginal pricing}
The theoretical underpinnings of locational marginal pricing trace
back to the spot-pricing framework of Schweppe~\textit{et
al.}~\cite{Schweppe1988}, which derived per-bus marginal energy
prices as the dual variables of a network-constrained dispatch
model.  Hogan~\cite{Hogan1992} formalised the LMP decomposition
\begin{equation}
\lambda_b = \lambda_{\mathrm{ref}} +
            \mathrm{LMC}_b + \mathrm{LCC}_b
\label{eq:lmp-decomp}
\end{equation}
into reference energy, marginal-loss component (LMC), and
marginal-congestion component (LCC).  This is the canonical
template that the present marginal-emission reconstruction adapts:
$\lambda_{\mathrm{ref}}$ is approximated by the
median-$\lambda$ of the electrically coherent island and the LMC
absorbs into a per-bus loss factor, while LCC is absorbed into the
island partition itself.

\subsection{Average versus marginal emission factors}
The distinction between average and marginal emission factors has
been debated since Hawkes~\cite{Hawkes2010} demonstrated that
marginal factors can differ by an order of magnitude from system
averages and that the choice materially affects build/no-build
decisions for energy-efficiency interventions.  Siler-Evans~\textit{et
al.}~\cite{SilerEvans2012} regressed historical EPA generator-level
data to fit MEFs at the regional Reliability Council level and
showed strong diurnal and seasonal variability.
Callaway~\textit{et al.}~\cite{Callaway2018} formalised the
distinction between \emph{operational} and \emph{decision-relevant}
emission factors, arguing that long-run capacity-induced effects
should be modelled separately from dispatch-time substitution
effects.  The present paper sits firmly in the operational /
short-run camp: the emission factor is computed against the
\emph{realised} dispatch published by CEN, not against a
counterfactual capacity scenario.

\subsection{ISO production tools}
PJM Interconnection's Locational Marginal Emissions
feed~\cite{PJMLME2024,PJMPrimer} publishes five-minute per-node
emission rates for CO$_2$, NO$_x$, and SO$_2$, decomposed using the
same energy/loss/congestion structure of the LMP itself.  CAISO's
\emph{Today's Outlook}~\cite{CAISOOutlook} surfaces a system-level
emission curve at five-minute resolution but is not nodal.
WattTime~\cite{WattTime2024} ships a balancing-area Marginal
Operating Emission Rate with a 72-hour forecast trained from
EIA-930 dispatch and EPA eGRID per-unit factors, but its output is
neither nodal nor unit-attributed.  NREL's
Cambium~\cite{NRELCambium} publishes \emph{long-run} marginal
factors projected through 2050 across \num{134} US balancing
zones — the natural counterpart for capacity-expansion studies as
opposed to operational dispatch.  Electricity
Maps~\cite{ElectricityMaps2024} publishes hourly per-zone
\emph{consumption-based} factors using flow-tracing and, since
2024, has discontinued its marginal feed citing fundamental
limitations of marginal accounting in scope-2
contexts~\cite{ElectricityMapsCritique}.

\subsection{Counterfactual / implicit-differentiation marginals}
A research-grade improvement on empirical correlation is the
counterfactual evaluation pioneered by
Valenzuela~\textit{et al.}~\cite{ValenzuelaASL2022}, who exploit
implicit differentiation of the QP/LP optimality conditions to
recover an exact dynamic locational marginal emission factor from a
solved dispatch.  A 2025 follow-up~\cite{Hong2025} argues that
counterfactual co-simulation should replace correlation-based MEFs
entirely.  The present pipeline is not yet counterfactual — it
remains an empirical reverse-engineering of the published
$\lambda^{\mathrm{CEN}}$ — but the open-source structure with the
PLEXOS \texttt{.accdb} oracle is designed so that the same
input set could be fed into the \texttt{gtopt} SDDP solver to
produce a counterfactual companion in future work.

\subsection{Prior CEN-specific work}
For Chile, Bonvallet~\textit{et al.}~\cite{Bonvallet2016}
co-optimised hydrothermal scheduling under non-convex irrigation
constraints and discussed the interpretation of dual variables on
hydraulic constraints as implicit water values.
Moreno~\textit{et al.}~\cite{Moreno2014} treat distributed-generation
deployment under uncertainty and quantify its abatement value
against a marginal-fuel benchmark.  The Coordinador's annual
\textit{Reporte Art.~72}~\cite{CENArt722024} aggregates emissions at
the system level, and the 4eChile guide~\cite{FourEChileGuide}
documents a manual marginal-emission reduction estimation tool for
Clean Development Mechanism (CDM) projects but neither delivers a
per-bus, per-quarter signal.  To the best of our knowledge no prior
work reconstructs $\varepsilon$ at locational and per-quarter
granularity for CEN strictly from the public CEN/SCVIC/PLEXOS
data triplet.

\subsection{Synthesis and methodological choice points}
Three choice points dominate the design space.  First, the
\emph{merit-stack picker}: should the marginal unit be the
last-dispatched cleared at $\lambda_{\mathrm{ref}}$
(``maximum CV~$\le \lambda$'')~\cite{SilerEvans2012}, or the
absolute-closest CV?  The pipeline picks the absolute-closest CV
because the closest-above is sometimes nearer than the
closest-below, particularly at low-CMG hours when biogas tiers sit
well below the gas tier.  Second, the \emph{forced-unit treatment}:
units flagged with \texttt{consigna}$\in$\{CI, MT, PMT, PP, PS,
PC\} are price-takers when at technical minimum, but when
operating interior they remain valid merit-margin candidates.
Third, the \emph{loss-factor decomposition}: a proper
PJM-style~\cite{PJMPrimer} decomposition would require the full
per-unit Power Transfer Distribution Factor matrix, which CEN does
not publish.  Instead the pipeline uses CEN's published
PLEXOS-solver-output MLF when available (extracted from the PID
\texttt{.accdb}) and falls back to the empirical
$\lambda_b/\lambda_{\mathrm{ref}}$ ratio when it is not.


% =====================================================================
% sota_extended.tex --- D1 --- extended state-of-the-art and standards
% review.  Appended at the END of sections/literature_review.tex by D6.
% =====================================================================

\subsection{Extended Review: Foundations, Standards, and Software}
\label{sec:literature-extended}

The reconstruction methodology presented in this paper sits at the
intersection of four research traditions: (i)~classical spot-pricing
theory, (ii)~the marginal-versus-average emission-factor literature,
(iii)~the recent industrialisation of hourly and locational
emission-rate datasets, and (iv)~the corresponding standardisation
landscape (ISO, GHG~Protocol, IEC, IEEE).  The remainder of this
subsection reviews each of these four threads, together with the
Latin-American institutional context and the open-source software prior
art.

\paragraph{Spot-pricing foundations.}
The locational marginal price (LMP) construct on which the present
work depends was developed in a sequence of papers by Caramanis, Bohn
and Schweppe.  Caramanis~\textit{et al.}~\cite{CaramanisBohn1982}
introduced optimal spot pricing as the dual of a generation-cost
minimisation, demonstrating that the per-bus shadow price internalises
losses, congestion, and reserve scarcity.  Bohn~\textit{et al.}~\cite{BohnCaramanis1984}
extended the construction to a continuous space-time setting and
proved the no-arbitrage equivalence with bilateral contract networks.
The textbook treatment of Schweppe~\textit{et al.}~\cite{SchweppeBook1988}
remains the canonical reference and was the basis for the Northeast
ISO market designs of the late 1990s.  Hogan~\cite{Hogan1992}
crystallised the contract-network formulation and the
energy/loss/congestion decomposition that
\cenpipe{} adopts as the structural template for its
$\lambda_b = \lambda_{\mathrm{ref}}\!\cdot\!\mathrm{MLF}_b$
correction.  None of these works contemplated emissions, but the
duality argument they articulate carries through verbatim once the
cost coefficient~$c_g$ in the primal is replaced by a fuel-emission
coefficient~$e_g$ (cf.~Section~\ref{sec:dual-formulation}).

\paragraph{Marginal versus average emission factors.}
The empirical-economics literature on marginal emission factors (MEFs)
opened with Hawkes~\cite{HawkesMEF2010}, who showed that for the GB
system the marginal $\mathrm{CO}_2$ rate exceeds the system average by
\SI{30}{\percent}--\SI{200}{\percent} depending on the demand level
and that abatement policies built on average factors systematically
under-credit demand-side measures.  Siler-Evans~\textit{et
al.}~\cite{SilerEvansMEF2012} regressed five years of EPA
continuous-emissions-monitoring data against ISO-level demand
fluctuations to obtain region-specific MEFs and exposed strong
diurnal and seasonal cycles.  Callaway~\textit{et
al.}~\cite{CallawayJAERE2018} formalised the operational vs.\
build-margin distinction and provided a welfare-theoretic argument
for using location-specific MEFs in policy design.  Holland~\textit{et
al.}~\cite{HollandMansurMuller2020} applied the same framework to the
post-2007 retirement of US coal capacity and quantified the resulting
shift of the marginal stack from coal to gas.  The present pipeline
contributes a new artefact to this body of work: a
counterfactual-free, per-bus, sub-hourly MEF reconstruction that uses
only the operator-published
$\lambda^{\mathrm{CEN}}$ and a static fuel-emission table.

\paragraph{Hourly and locational emission datasets.}
Three production-grade datasets dominate the hourly-emission space.
Hawkins~\cite{HawkinsPJM2024} reports on the operational
characteristics of PJM Interconnection's locational marginal emissions
feed after one full year of operation and confirms that hourly nodal
factors track the marginal-fuel composition with sub-\SI{5}{\percent}
median residuals.  The Tomorrow / Electricity Maps open
dataset~\cite{ElectricityMapDataset} publishes hourly per-zone
consumption-based factors using flow-tracing across European,
American, and Australasian balancing areas.  WattTime's Marginal
Operating Emission Rate~\cite{WattTimeMOER} is balancing-area
resolved and explicitly marginal but not nodal.  Singularity
Energy~\cite{SingularityHourly} delivers an hourly attribution layer
oriented towards corporate Scope-2 disclosure.  None of these vendors
publishes Chilean coverage, which motivates the present work.

\paragraph{Standards.}
The relevant standardisation environment spans three families:
\begin{itemize}
  \item ISO greenhouse-gas accounting:
        ISO~14064-1~\cite{ISO14064_1} (organisation-level inventory),
        ISO~14064-2~\cite{ISO14064_2} (project-level emission
        reductions),
        ISO~14064-3~\cite{ISO14064_3} (verification),
        and
        ISO~14067~\cite{ISO14067} (product carbon footprint).  The
        \cenpipe{} pipeline supplies the
        location-and-time-resolved emission factor required by
        clauses 7.2.2 and 8.2.3 of ISO~14064-1 for indirect
        emission inventory at sub-annual granularity.
  \item Corporate accounting: the WRI/WBCSD GHG~Protocol Scope-2
        Guidance~\cite{GHGPScope2} mandates a dual reporting
        framework (location-based + market-based).  RE100~\cite{RE100Hourly}
        and the 24/7 Carbon-Free Energy
        Compact~\cite{CFE247Compact} represent the strict end of the
        spectrum, requiring hour-by-hour matching of consumption and
        contracted carbon-free supply.  The
        \cenpipe{} 15-minute output natively satisfies the
        granularity requirement of both schemes.
  \item Power-system data exchange: IEC~61968-13~\cite{IEC61968_13}
        defines the Common Information Model (CIM) profile for
        electricity-market data, and IEC~61970/CGMES~\cite{IEC61970_CGMES}
        is the European common-grid-model serialisation that ENTSO-E
        ISOs use for cross-border data exchange.  IEEE~Std
        1547-2018~\cite{IEEE1547_2018} governs distributed energy
        resource interconnection and is directly relevant to the BESS
        case study of Section~\ref{sec:bess-case}, while
        IEEE~Std~2030.5~\cite{IEEE2030_5} specifies the smart-energy
        application protocol.  The Coordinador El\'ectrico Nacional
        Norma T\'ecnica de Servicios
        Complementarios~\cite{CENNT_SSCC2024} is the Chilean
        equivalent that frames the storage-and-reserve products under
        which a battery operator would monetise the dispatch
        trajectories computed in
        Section~\ref{sec:bess-case}.
\end{itemize}

\paragraph{Latin-American context.}
The Chilean academic literature on dispatch optimisation is
relatively recent and centred on a small number of groups.
Bonvallet~\textit{et al.}~\cite{Bonvallet2016} treated hydrothermal
scheduling with non-convex irrigation constraints and quantified the
implicit water values that drive marginal-cost rotation in the
Maule and Laja basins.  Mocarquer and Sauma~\cite{MocarquerSauma2018}
documented the post-2014 market reforms and their effect on
investment incentives.  At the policy level, the Chilean Ministry
of Energy roadmap to carbon neutrality
2050~\cite{ChileMinEnergia2050} sets the macroscopic abatement
trajectory, and the 2024 CEN dispatch-decarbonisation
roadmap~\cite{CENRoadmap2024} translates that target into operational
milestones for coal phase-out, storage build-out, and ancillary
services reform.  None of these documents publishes the
per-bus per-quarter marginal-emission signal that
\cenpipe{} provides; they all rely on annual or
monthly aggregates derived from PLEXOS post-processing.

\paragraph{Software prior art.}
Three open-source projects implement related capabilities, although
none operates on Chilean data.  PyPSA-Eur~\cite{PyPSAEurEmissions} ships
an emissions module that integrates IPCC AR6 fuel factors with the
solved dispatch to produce per-snapshot per-node emission rates;
unlike \cenpipe{} it requires the LP to be re-solved
rather than reconstructing $\varepsilon$ from a published $\lambda$.
GenX~\cite{GenXCarbon} provides a cap-and-trade extension that
endogenises the carbon price as the dual of an aggregate emission
constraint, conceptually parallel to the joint LP of
Section~\ref{sec:dual-formulation}.  SDDP.jl~\cite{SDDPjlCarbon}
includes a tutorial on adding a carbon constraint to a multi-stage
hydrothermal stochastic program; \texttt{gtopt} aims to support the
same construction natively in C++.  Operationally, the Electricity
Maps dataset~\cite{ElectricityMapDataset} is the closest commercial
analogue: hourly consumption-based factors from real-time data
ingestion, but with continental rather than sub-system-level
resolution and with no Chilean coverage.
